$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-04-29 paper/paper.tex
Go through the typign rules again

In the \picalc{} both scope restriction and input introduce abstractions.
The bodies of these constructs need a way of referring to their argument.
Using names as variable references is a popular option amongst humans.
However, names are cumbersome to mechanize: inserting a new variable into a context means proving that the name of such variable is different to all other variable names in context.
Moreover, names are of no significance to machines.

The idea de Bruijn had \cite{} was to use a natural number $n$ (aka \emph{index}) to refer to the variable introduced $n$ binders ago --- and binders no longer introduce names.
\begin{example}
\hfill{}\\
The \picalc{} term \hfill{} $(\textbf{new} \; x) (x ( y ) . y \langle \texttt{"Hello world!"} \rangle . \mathbf{0} \parallel (\textbf{new} \; y) (x \langle y \rangle . y ( z ) .\mathbf{0}))$ \\ is represented as \hfill{} $(\textbf{new}) (0 () . 0 \langle \texttt{"Hello world"} \rangle . \mathbf{0} \parallel (\textbf{new}) (1 \langle 0 \rangle . 0 () . \mathbf{0}))$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.
Humans may find this confusing.
There is however no reason not to annotate these indices with the original names bestowed upon them by humans.
Machines can then manipulate references mechanically whilst using names to present terms to humans.

@@

\begin{note}
  We will often work with input and output multiplicities as pairs.
  We use the notation $\Carrier^2$ to stand for the type $\Carrier \times \Carrier$, and $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where all properties are trivially lifted.
\end{note}

@@

We introduce two judgments: one for variable references (Definition \ref{def:varref}), one for processes (Definition \ref{def:types}).

@@

  Like processes, variable reference judgments (Figure \ref{fig:variable-references}) are indexed by a typing context and an input and output usage context.
  In addition to these, variable reference judgments are also indexed by their position in the contexts, their type, and their usage annotation.

@@

  Our judgment on processes is of the form $\types{\gamma}{\Gamma}{P}{\Delta}$ (Figure \ref{fig:types}).
  This assertion combines a judgment on types ($\gamma$) and a judgment on usages ($\Gamma$ and $\Delta$).
  They could be kept as two separate judgments, but doing so would risk incoherence: if there is any choice to be made, the judgment on types could choose to go left while the judgment on usages chooses to go right.
  The constructions instroduced in this paper make the typing derivations completely syntax directed and thus this problem inexistent, but we nevertheless choose to keep both judgments as one for simplicity.

  The typing relation \textsc{Types} presented in Figure \ref{fig:types} is indexed over a context of types $\gamma$, a context of input multiplicities $\Gamma$, the process $P$ that is being typed, and the context of output capabilities $\Delta$, and notated as $\types{\gamma}{\Gamma}{P}{\Delta}$.
  \todo{empty output context}
  The typing rules are in one-to-one correspondence with the underlying syntax for processes, and thus we say they are syntax-directed.

  Process termination is typed under any typing context, and preserves the multiplicity context unchanged.

  Scope restriction accepts three parameters: the type $t$ of data sent over the channel; the input and output multiplicities $x$ of the data sent over the channel; and the multiplicity $y$ of the channel itself, which is used both as input and as output.
  The typing derivation of the continuation process $P$ has a channel of data type $t$ and data multiplicity $x$ added to the typing context, and the multiplicities $(y , y)$ added to the input multiplicity context.
  The continuation must completely exhaust this channel, to that end we substitute $(y , y)$ by $(\zero , zero)$ in the output context.
  \todo{comment on balancing in chan rule}

  Input requires a proof $chan_i$ locating a channel with data type $t$, data multiplicities $x$ and multiplicity $\ell_i$ \todo{} at index $i$ in the typing and multiplicity contexts $\gamma$ and $\Gamma$.
  The result of exhausting $\ell_i$ is the output multiplicity $\Xi$.
  The data type $t$ is then used to expand the typing context of the continuation $cont$, while the data multiplicities $x$ are appended to the output context $\Xi$ to form the input context over which $P$ is typed.
  These multiplicities $x$ must be exhausted by typing derivation of the continuation process.
  Note that the index $i$ is then used as part of the syntax.

  \todo{note that every new multiplicity introduced in an input context must be exhausted in the output context}

  Similar to input, output requires a proof $chan_i$ locating a channel with data type $t$ and data multiplicities $x$ in $\gamma$ at index $i$, however this time the channel must have multiplicity $\ell_o$.
  The result of exhausting $\ell_o$ is the output multiplicity $\Delta$.
  That output context $\Delta$ must have a variable of type $t$ and multiplicities $x$ at position $j$.
  This second proof represents the data that is sent over the channel.
  The result of exhausting the multiplicities $x$ of that variable is the context $\Xi$.
  The continuation process must be typed under this context $\Xi$.

  Finally, parallel composition requires its two subprocesses to be typed such that the output context of the first is the input context of the second.

>>>>

In the \picalc{} both scope restriction and input introduce abstractions.
The bodies of these constructs need a way of referring to their argument.
Using names as variable references is a popular option amongst humans.
However, names are cumbersome to mechanize: inserting a new variable into a context means proving that the name of such variable is different to all other variable names in context.
Moreover, names are of no significance to machines whatsoever.

The idea de Bruijn had \cite{} was to use a natural number $n$ (aka \emph{index}) to refer to the variable introduced $n$ binders ago --- and to have binders no longer introduce names.
\begin{example}
\hfill{}\\
The \picalc{} term \hfill{} $(\textbf{new} \; x) (x ( y ) . y \langle \texttt{"Hello world!"} \rangle . \mathbf{0} \parallel (\textbf{new} \; y) (x \langle y \rangle . y ( z ) .\mathbf{0}))$ \\ is represented as \hfill{} $(\textbf{new}) (0 () . 0 \langle \texttt{"Hello world!"} \rangle . \mathbf{0} \parallel (\textbf{new}) (1 \langle 0 \rangle . 0 () . \mathbf{0}))$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.
Humans may find this confusing.
There is however no reason not to annotate these indices with the original names bestowed upon them by humans.
Machines can then manipulate references mechanically whilst using names to present terms to humans.

@@

\begin{note}
  We will often work with input and output multiplicities as pairs.
  We use the notation $\Carrier^2$ to stand for a $\Carrier \times \Carrier$ tuple, and $\opsquared{x}{y}{z}$ to stand for a monoidal operation on pairs of multiplicities -- where all properties are trivially lifted.
\end{note}

@@

We introduce two judgments: one for variable references (Definition \ref{def:varref}), one for processes (Definition \ref{def:types}).
Both of them combine a judgment on types and a judgment on usages.
These could be kept separate, but doing so would risk incoherence: if typing derivations had any choice to make, the judgment on types could choose to do one thing while the judgment on usages chooses to do another.
The constructions instroduced in this paper make the typing derivations completely syntax directed and render this problem inexistent, but we nevertheless choose to keep both judgments as one, for simplicity.

@@

  The \textsc{VarRef} relation presented in Figure \ref{fig:variable-references} is indexed by a typing context and an input and output usage context.
  In addition to these, it also keeps track of its position in the contexts, its type, and its usage annotation.

@@

  The typing relation \textsc{Types} presented in Figure \ref{fig:types} is indexed by a typing context, an input and an output usage context, and the process $P$ being typed.

  The process termination case is only concerned with usage contexts.
  Its input usage context is returned as leftovers.

  Scope restriction expects three arguments: the type $t$ of data being transmitted; the input and output multiplicities $x$ of the data being transmitted; and the multiplicity $y$ assigned to the channel itself.
  The same multiplicity $y$ is used as the input multiplicity and as the output multiplicity of the channel, so these are balanced.
  The typing derivation of the continuation process $P$ is provided with this channel with multiplicity $(y \comma y)$.
  Process $P$ must then completely exhaust this channel, and to that end we substitute $(y , y)$ in the input context with $\ell_{\o}$ in the output context.

  \todo{comment on synonyms $\ell_i$ etc}

  Receiving requires a proof $chan_i$ that selects a channel at index $i$ with multiplicity $\ell_i$ available, over which data with type $t$ and multiplicity $x$ is going to be sent.
  Note that this index $i$ is then used in the syntax of the typed process.
  After consuming $\ell_i$ from this channel, we use the leftover context $\Xi$ to type the continuation.
  The continuation is also provided with the received element --- of type $t$ and multiplicity $x$ at index $\constr{zero}$.
  This multiplicity $x$ must be completely exhausted by the continuation process.

  Similar to input, output requires a proof $chan_i$ that selects a channel at index $i$ with multiplicity $\ell_o$ available, over which data with type $t$ and multiplicity $x$ is going to be sent.
  Note that this index $i$ is then used in the syntax of the typed process.
  After consuming $\ell_o$ from this channel, we use the leftover context $\Delta$ to type the continuation.
  The continuation must select the element at index $j$ that is being sent.
  This element must have type $t$ and at least multiplicity $x$ --- as per the channel's type.
  The result of consuming these multiplicity $x$ is the leftover context $\Xi$, which is then used to type the continuation process.
  Note that both indices $i$ and $j$ are used in the syntax of the typed process.

  Finally, parallel composition is only concerned with usage contexts.
  The leftovers of the left-hand process are used to type the right-hand process.
